---
title: "PA1_template.Rmd"
author: "Elena K"
date: "26/04/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reproducible Research: Peer Review Project 1

This is peer review project 1 for the course **Reproducible Research**

What needs to be done:

 * Code for reading in the dataset and/or processing the data
 * Histogram of the total number of steps taken each day
 * Mean and median number of steps taken each day
 * Time series plot of the average number of steps taken
 * The 5-minute interval that, on average, contains the maximum number of steps
 * Code to describe and show a strategy for imputing missing data
 * Histogram of the total number of steps taken each day after missing values are imputed
 * Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
* All of the R code needed to reproduce the results (numbers, plots, etc.) in the report

## Code for reading dataset

```{r Libraries_and_data, echo=TRUE}
library(tidyverse)  #helps wrangle data
library(lubridate)  #helps wrangle date attributes
library(ggplot2)  #helps visualize data

## Task 1, read data
##steps <- read.csv("~/../Desktop/ProjectData/activity.csv")

url <-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
#define two temporary files
temp <- tempfile()
temp2 <- tempfile()

# Download and unzip data
download.file(url, temp)
unzip(zipfile = temp, exdir = temp2)
steps <- read.csv(file.path(temp2, "activity.csv"))

# Remove temporary files
unlink(c(temp, temp2))

```

## Code for Processing Data

```{r process_data, echo=TRUE}
steps$dt <-
  parse_date_time(#"23/8/22 12:45", "dmy HM", tz="")
    paste(steps$date, " ", as.character(steps$interval%/%100),":",as.character(steps$interval%%100),sep="")
    ,"%Y-%m-%d %H:%M", tz="")         
steps$date <- as.Date(steps$date)
steps$hour <- steps$interval%/%100
steps$minutes<-steps$interval%%100

summary(steps)
table(is.na(steps$steps))
```

## Histogram of Daily Steps

```{r daily_steps, echo=FALSE}
daily_steps <- steps %>% filter(!is.na(steps)) %>%
  group_by(date) %>% summarise(steps_per_day = sum(steps))

hist(daily_steps$steps_per_day, breaks=10)
```

## Average and Median Steps taken every Day
```{r ave_steps, echo=FALSE}
steps %>% filter(!is.na(steps)) %>%
  group_by(date) %>% 
  summarise(steps_per_day = sum(steps)) %>%
  ungroup() %>% 
  summarize(average_steps_per_day = mean(steps_per_day), 
                                         median_steps_per_day=median(steps_per_day))
```

## Time series plot of the average number of steps taken

```{r steps_taken, echo=FALSE}
steps %>% filter(!is.na(steps)) %>%
  group_by(date) %>% 
  summarise(steps_per_day = sum(steps)) %>%
  ggplot() +
  geom_col(mapping = aes(x=date ,y=steps_per_day), fill="cyan")+
  geom_smooth(mapping = aes(x=date ,y=steps_per_day))
```

## The 5-minute interval that, on average, contains the maximum number of steps
```{r, echo=FALSE}

interval_with_max_steps <- (steps %>% filter(!is.na(steps)) %>%
  group_by(interval) %>% summarize(average_per_interval = mean(steps)) %>%
  filter(average_per_interval==max(average_per_interval)))$interval

steps %>% filter(!is.na(steps)) %>%
  group_by(interval) %>% summarize(average_per_interval = mean(steps)) %>%
  ggplot() +
  geom_point(mapping = aes(x=interval,y=average_per_interval), color="blue") +
  geom_vline(aes(xintercept=interval_with_max_steps, color="red")) +
  annotate("text", x=1500, y=200,
           label=paste0("Interval for Max average steps: ",interval_with_max_steps), 
          color="red")
  

```


## Code to describe and show a strategy for imputing missing data

I choose to populate data with the number of steps for the "typical day":

```{r, echo=TRUE}
ave_steps_per_interval <-steps %>% filter(!is.na(steps)) %>%
  group_by(interval) %>% summarize(average_per_interval = mean(steps))

steps_v2 <-steps
steps_v2$steps[is.na(steps$steps)]<-
  ave_steps_per_interval$average_per_interval[match(steps_v2$interval,ave_steps_per_interval$interval)][which(is.na(steps_v2$steps))]

summary(steps_v2)
 
```


## Histogram of the total number of steps taken each day after missing values are imputed
```{r, echo=FALSE}
daily_steps_v2 <- steps_v2 %>% filter(!is.na(steps)) %>%
  group_by(date) %>% summarise(steps_per_day = sum(steps))

hist(daily_steps_v2$steps_per_day, breaks=10)
```


## Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r, echo=FALSE}
weekdays1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
steps_v2$wday <- factor((weekdays(steps_v2$date) %in% weekdays1), 
                   levels=c(FALSE, TRUE), labels=c('weekend', 'weekday'))
steps_v2 %>% filter(!is.na(steps)) %>%
  group_by(wday, interval,) %>% summarize(average_steps = mean(steps)) %>%
  ggplot() +
  geom_col(mapping = aes(x=interval,y=average_steps, fill=wday)) +
  #legend_position="bottom")+
  facet_wrap(~wday)
```





